---
title: "Lab 05 - La Quinta is Spanish for next to Denny's, Pt. 2"
author: "Holland Sun"
date: "2026-Feb"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1

```{r}
dn_ak <- dennys %>%
  filter(state == "AK")

lq_ak <- laquinta %>%
  filter(state == "AK")
```
There is ```r nrow(dn_ak)``` Denny's in Alaska

There is ```r nrow(lq_ak)```  La Quinta in Alaska

### Exercise 2
So there is ```r nrow(dn_ak) * nrow(lq_ak)``` parings Between these locations.

### Exercise 3
```{r,fulljoinak ,warning=FALSE}
dn_lq_ak <- full_join(dn_ak, lq_ak,
  by = "state"
)
```

### Exercise 4
There is ```r nrow(dn_lq_ak)```  obserrvations in the ```dn_lq_ak``` data frram.

The names of the variables are ```r names(dn_lq_ak)```

### Exercise 5
I think is ```mutate``` function that
 can add a new variable to a data fram while keeping the existing varriables in tidyverse.

### Exercise 6

```{r}
source("haversine.R")
dn_lq_ak <- dn_lq_ak %>%
  mutate(distance = haversine(longitude.x, latitude.x,
                              longitude.y, latitude.y, round = 3))
```

### Exercise 7
```{r}
dn_lq_ak <- dn_lq_ak %>%
  group_by(address.x) %>%
  mutate(min_distance = min(distance)) %>%
  ungroup()
```

### Exercise 8
```{r}
dn_distance_MIN <- dn_lq_ak %>%
  group_by(address.x) %>% 
  summarise(min_distance = min(distance))

ggplot(dn_distance_MIN, aes(x = min_distance)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(x = "Minimum Distance  (km)",
       y = "Density") +
  theme_minimal()

summary(dn_distance_MIN$min_distance)
```

<span style="color: deepskyblue;"> *(I thought maybe boxplot also can be a good idea, but density looks much better, except the later one may provides some "overfiting/misleading" trend since we just have 3 data point)* </span>

### Exercise 9,10,11
Since we are performing some kind of repetitive analysis, why not create a function? 
and We could even analyze all states at once
```{r}
PLmin_dis_cal <- function(state_index) {
  dn_state <- dennys %>% filter(state == state_index)
  lq_state <- laquinta %>% filter(state == state_index)
  
  dn_lq <- full_join(dn_state, lq_state, by = "state")
  
  dn_lq <- dn_lq %>%
    mutate(distance = haversine(longitude.x, latitude.x,
                                longitude.y, latitude.y, round = 3))
  
  mindist <- dn_lq %>%
    group_by(address.x) %>%
    summarise(min_distance = min(distance))
  
  return(mindist)
}

```

#### For NC

```{r,nc cal,warning=FALSE}

nc_mindist <- PLmin_dis_cal("NC")

ggplot(nc_mindist, aes(x = min_distance)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(x = "Minimum Distance (km)", y = "Density",
       title = "NC") +
  theme_minimal()

summary(nc_mindist$min_distance)

```

#### For TX
```{r,tx cal,warning=FALSE}

tx_mindist <- PLmin_dis_cal("TX")

ggplot(tx_mindist, aes(x = min_distance)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(x = "Minimum Distance (km)", y = "Density",
       title = "TX") +
  theme_minimal()

summary(tx_mindist$min_distance)

```

#### For NY
<span style="color: deepskyblue;"> Because my favorite comic I used to read in high school geography class was Spider-Man, I chose New York lol  </span>
```{r,ny cal,warning=FALSE}

ny_mindist <- PLmin_dis_cal("NY")

ggplot(ny_mindist, aes(x = min_distance)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(x = "Minimum Distance (km)", y = "Density",
       title = "NY") +
  theme_minimal()

summary(ny_mindist$min_distance)

```


### Exercise 12
Among the states examined, Mitch Hedberg's joke is most likely to hold true in Texas. 

<span style="color: deepskyblue;"> Although I didn't realized this advantage when I initially chose the density plot, I'd like to shamelessly point out this kind of intuitive visualization of distributions makes it easier for us to draw conclusions at a glance is some
question like here. Moreover, we are essentially measuring the entire population here which makes the density plot an even more faithful representation.  </span>

The distribution of minimum distances in TX is heavily right-skewed with a median of only 3.37 km and a mean of 5.79 km. In contrast, NC ( median = 53.46 mean= 65.44) and NY (median = 24.16 mean=33.58) show much larger distances. Although Alaska has a small mean distance (4.41 km), the sample size of only 3 Denny's locations makes this unreliable.

### Exercise 14
<span style="color: deepskyblue;"> Ok, here's another Holland's comments transcending time: </span>

<span style="color: deepskyblue;">*As I mentioned in my email this week, I initially tried to complete a different lab. In that earlier lab (I can't remember if it was Lab 05, 06, or 07 I can never remember these numbers) . I vaguely recall there was a question about an ugly or sad plot where you could propose improvements, even if you couldn't actually implement them by code. In that exercise, I mentioned wanting to create a heatmap. Seeing your ridgeline plot here inspired me: I could build a heatmap on top of this, though it would sacrifice some information (the shape of the distribution). The idea is to use the mean distance as a representative value and assign different colors to different states.*</span>
 
```{r}
# Source pattern: https://stackoverflow.com/a/36149654
# Posted by inscaven
# Retrieved 2026-02-08, License - CC BY-SA 3.0

library(usmap) #import the package
library(tidyverse)
library(usmap)
library(ggplot2)
library(sf)
library(grid)

# 1) Closest La Quinta distance for each Denny's (all states)
dn_lq_full <- full_join(
  dennys, laquinta,
  by = "state",
  suffix = c(".d", ".l")
) %>%
  filter(!is.na(latitude.d) & !is.na(latitude.l)) %>%
   filter(!state %in% c("AK", "HI")) 

dn_lq_full_mindist <- dn_lq_full %>%
  group_by(address.d) %>%
  mutate(distance = haversine(longitude.d, latitude.d, longitude.l, latitude.l)) %>%
  summarize(
    closest = min(distance, na.rm = TRUE),
    state   = first(state),
    .groups = "drop"
  )

# make some adjustment
state_mean <- dn_lq_full_mindist %>%
  group_by(state) %>%
  summarise(mean_closest = mean(closest, na.rm = TRUE))
states_sf <- usmap::us_map(regions = "states")
states_sf <- states_sf %>%
  left_join(state_mean, by = c("abbr" = "state"))

ggplot(states_sf) +
  geom_sf(aes(fill = mean_closest), color = "gray60", linewidth = 0.4) +
  scale_fill_viridis_c(
    name = "Mean Min_Distance (km)",
    option = "plasma",
    direction = -1,    
    na.value = "gray90"
  ) +
  theme_void() +
  scale_fill_viridis_c(option = "viridis")+
  labs(title = "Average Minimum Distance from Denny's to  La Quinta by State")
```


See you next lab :)
